Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-06-17 Manuscript/SupplementaryMaterial/TreeBuilding.tex
Supplementary Materials init

%HEADER
\documentclass[12pt,letterpaper]{article}

%packages
% NC: Just to flag here how I'll comment on LaTeX files. I'll use % and then NC. Of course in git diff it'll all show up as an addition.
% NC: You could add all the packages etc to a template .sty file to save writing it all up here. Rich's ms.sty that I provided on DropBox is a good example
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{amsfonts}
\usepackage{verbatim}
\usepackage[english]{babel}
\usepackage{pifont}
\usepackage{color}
\usepackage{setspace}
\usepackage{lscape}
\usepackage{indentfirst}
\usepackage[normalem]{ulem}
\usepackage{booktabs}
%\usepackage{nag}
\usepackage{natbib}
%\usepackage{bibtex}
\usepackage{float}
\usepackage{latexsym}
%\usepackage{hyperref} 
\usepackage{url}
%\usepackage{html}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
%\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{amstext} % Add and remove packages as necessary for your manuscript.

% NC: Added by NC
\usepackage{enumerate}
\usepackage[osf]{mathpazo} % palatino is a nicer font! 
\pagenumbering{arabic}	% Page numbers

%Pagination style and stuff
\linespread{1.66} % All text should be double-spaced with occasional exceptions for tables. 
\raggedright
\setlength{\parindent}{0.5in}

\setcounter{secnumdepth}{0} % Our sections are not numbered and our papers do not have Tables of Contents. We don't present a list of figures or list of tables, either. Any common font is fine. (A common sans-serif font should be used on figures, but figures should be separate from the LaTeX document.)

%\pagestyle{empty}

\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}

\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}

\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}

\renewcommand{\tableofcontents}{}

\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib

\begin{document}

%For a colour version of the draft to a search>replace '-BW'>'' 
% %\modulolinenumbers[1] 	% Line numbering on every line

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: Missing data and topology in Total Evidence matrices %RH = Running Head

\bigskip
\medskip
\begin{center}

%Title
\noindent{\Large \bf Effect of missing data in matrices containing living and fossil taxa on topological accuracy}
%Alternative punchier title: "Bayesian Total Evidence methods fail to recover correct topologies from matrices containing both living and fossil taxa"
%Alternative punchier title: "Maximum Likelihood outperforms Bayesian methods for inferring topology from Total Evidence matrices containing both living and fossil taxa"

% NC: Not sure about any of these. I actually think I prefer your talk title. So something like: The effects of missing data on the ability of Total Evidence approaches to recover correct tree topologies. No rush to decide. But needs Total Evidence in there somewhere.
\bigskip

%Authors
\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$, other authors $^3$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland;\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland;\\
$^3$Somewhere else }\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Thomas Guillerme, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
\vspace{1in}
% NC: It's not called the Republic of, just Ireland!

\newpage
\begin{abstract}
% NC: I'm not going to give this much work as it'll change before we submit anyway. Note though that and abstract must NOT just repeat stuff in the intro. Needs rephrasing at the very least.
Living species represent less than 1\% of all species that have ever lived. % NC: I was chatting to Andy Purvis about this the other day and he said lots of palaeo folk hate that stat. So let's not start the abstract with it!
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology.
This fact has led to an increasing consensus among scientists that both living and fossil taxa must be included in macroevolutionary studies.
One approach, the Total Evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both living and fossil taxa at the tips.
Although the Total Evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

In this study we assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the matrix:
(1) the proportion of living taxa with no morphological data,
(2) the amount of missing data in the fossil record, and
(3) the overall number of morphological characters in the matrix.
We find that, in a Bayesian framework, difficulties in recovering a stable topology are mainly driven by the missing data in the molecular part of the matrix (for which fossil taxa have no data).
In a Maximum Likelihood framework, however, topology is not directly affected by missing data \textit{per se}, but by the number of morphological characters shared among the taxa.
Therefore, the two main drivers of incorrect topologies are the overall number of morphological characters and the number of living species with no morphological data.

Our results suggest that, in order to use Total Evidence approaches, one should reduce the missing data in the morphological part of the matrix for living species and use a Maximum Likelihood framework to fix the topology prior to the overall Bayesian phylogenetic inference process.
%More results!
\end{abstract}

\noindent (Keywords: missing data, Total Evidence, Bayesian, Maximum Likelihood, topology)\\

\vspace{1.5in}

\newpage

\section{Introduction}

Although most species that have ever lived are now extinct % NC: This avoids giving the dodgy stat of 1%
    \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}. %or is citing of your own papers in the too "douchebagy" for the second sentence?
% NC: Yeah be careful with self citing. Fine if it's a good fit, but it's totally not here!!! A better citation would be a big biodiversity study, e.g. the bird phylogeny paper.
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}.
This has led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}.
However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013}). %Not a nice sentence

Up to now, three main approaches have been used to place both living and fossil taxa % NC: BE consistent. Either living and fossil, or fossil and living. - TG: Let's make it living and fossil then.
    into phylogenies.
These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). % NC: Not sure about bit in brackets - TG: it's the idea of fossil occurence (age) only when node and fossil occurence + topology when tip.
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny.
Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty (but see \citealp{spencerefficacy2013}).
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence age % NC: Need to check the exact phrase needed here. Is it range of occurrence? Occurrence dates? etc. - TG: in paleo jargon we have FAD (first apparition datum) and FOD (first occurrence datum). FAD is always hypothetical (we can't really exactly know when the taxa appeared) and FOD is always subjective (it depends when the fossil was found, maybe next year someone will find another one in older layers). So for the dating, because it's relaxed clock, people use FOD (and also probably because they don't care since it doesn't make a huge difference I guess).
 are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013}.
However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa. %There might be a philosophico-semantic distinction to do here. Is TEM a method or an approach? I'll say that it is an approach because it's more like an idea (put every thing together) than a way to do (how to put everything together). It makes more sense to me compared to a real method e.g. a probabilistic method such as ML where the method states precisely how to do stuff.
% NC: Tricky. Don't think it's too important though. 
% NC: I do think Total Evidence needs capital letters though, like ML and Bayesian - TG: fixed
This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{eernissetaxonomic1993}.
Total Evidence methods have been successfully applied to empirical data \citep{pyrondivergence2011,ronquista2012,schragocombining2013}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies.
However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data.
%In particular it requires morphological data from both living and fossil taxa, both of which are known to be scarce. % NC: Really? I know fossil data is known to be scarce, but living data? I think people would be surprised. And they are surprised. We were surprised! Needs a citation either way.
%TG: how about this:
The morphological data for living taxa is rarely collected when molecular data is available (e.g. \citealp{O'Leary08022013} vs. \citealp{meredithimpacts2011}), and for fossil taxa, the scarcity of the fossil record only allow to collect the data available (for example, in vertebrates, the solidest parts of the skeleton \citeal{sansomfossilization2013}).
Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}. % Trevor: Missing data also influences support values that trend to decrease

% NC: You need to take a bit of time to learn about singular and plural grammar in English. It's something you always do wrong. e.g. The Total Evidence method *uses* this and *its* parameters are as follows. Versus: Total Evidence methods *use* this and *their* parameters are as follows. Sive and I won't be around to correct your English forever. - TG: my French demons are back!

%Paused here - 15/063

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013}.
Missing molecular data has been seen by some authors as an issue because it can decrease phylogenetic signal (i.e. % define phylogenetic signal here) 
in some parts of the tree, especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by:
(i) including a "modest" number of highly-covered % what do you mean by highly-covered?
genes (i.e. approximatively half of the genes; \citealt{rouresite-specific2011});
(ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and
(iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}.
Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013}.
Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies \citep{sansomfossilization2013}.
However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{wiensmissing2003}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point.

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil species. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference from Total Evidence methods.

% NC: Somewhere here you need to flag that you're only looking at topologies and briefly explain why.

Here we use simulations to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. The molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}. The effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}, therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness of the morphological part of the matrix:
\begin{enumerate}
\item the proportion of living taxa with no morphological data;
\item the amount of missing data in the fossil taxa; and
\item the overall number of morphological characters for both living and fossil taxa in the matrix.
\end{enumerate}
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We chose these parameters because they reflect real-life biases in data availability. The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis. Missing data in fossil taxa is very common due to preservation biases (citation?), and 
% OK I'm not sure what the "real-life" reason for three is.
%This paragraph is crap
% NC: I think it's fine!

We find that when using a Maximum Likelihood approach, as missing data increases, the likelihood of recovering the correct tree topology decreases. However, even with no missing data, Total Evidence matrices dramatically reduce the performance of Bayesian methods for inferring tree topology. We propose that this drastic difference between Bayesian and Maximum Likelihood methods is due to a flattening of the likelihood landscape caused by the unavoidable amount of missing molecular data for fossil taxa in a Total Evidence matrix. We make suggestions for how best to deal with this issue when inferring phylogenies from Total Evidence matrices.

% NC: You don't want to tell the whole story here. Just a results summary and a quick line of interpretation

%This problem leads to an inability of Bayesian chains to pick up the sub and near optimal topologies leading to high variance in the posterior tree distribution and really low support in the consensus tree.
%These results should encourage authors in not using Bayesian inference to explore the topological landscape of their phylogeny.
%Instead, we suggest that authors should first find the Maximum Likelihood tree and use it's topology as a fix prior in their Bayesian inference.
%We also propose a new method for comparing Bayesian posterior tree distribution's topology available in a wonderful R package. %Or am I just dreaming a bit too much here?

% NC: Dreaming a bit much I think. There's a discussion to be had here about whether it's really a *new* method and whether you really want to make an R package per se. We definitely want to share the code

%-----------
% NC: It might look like I've done a lot to this but actually I think it's nearly there. A good thing to take from this is that writing a paper is *hard*. It gets easier, but it's still a deeply involved process with lots of writing and re-writing. Introductions are particularly difficult because if you don't set the scene, it's hard to convince people to get on board for the rest of the paper. This is why I said there was no way this was being submitted by the end of June! Still a long way to go I'm afraid in terms of time and effort. But that doesn't mean you're not doing well, *every* project is like this. 2 years is about average from conception to sending out a manuscript, and you only started work on this last spring.
%------------

\newpage
\section{Methods} %no material needed?
% NC: Depends on the journal. See what Syst Biol asks for as section titles.
To explore how missing data in Total Evidence matrices influences tree topology we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}). %NC: does Syst Biol use Fig. or Figure?
\begin{enumerate}
\item{Generating the matrix} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the "true" tree; Table ~\ref{Tab_glossary}) and used it to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix; Table ~\ref{Tab_glossary}).
\item{Removing data} \label{step:remove_data} \\
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of missing living taxa ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$) % NC: make sure to use the same descriptions here and in the introduction 
(the resulting matrices are called hereafter "missing-data" matrices; Table ~\ref{Tab_glossary}).
\item{Building phylogenies} \label{step:build_phylo} \\
We built phylogenetic trees from the "complete" matrix and from the "missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the "best" tree; Table ~\ref{Tab_glossary}) and multiple trees inferred from matrices with missing morphological data (hereafter called the "missing-data" trees; Table ~\ref{Tab_glossary}). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies} \label{step:compare_topo} \\
We compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies.
\end{enumerate}
To measure the effect of missing data distribution, we repeated steps ~\ref{step:generate_matrix} to ~\ref{step:compare_topo} with the exact same fixed parameters 51 (3$\times$17) times. %that's because I had 3 cluster accounts available... If it's a too odd number, no worries, I'll just remove the last chain.
A list of all the terms used in this paper is available in table ~\ref{Tab_glossary}.

\begin{figure}
\centering
\includegraphics[keepaspectratio=true]{Figures/TEM_Fig_outline-BW.pdf}
\caption{Protocol outline.
(1) We generated a random tree (the "true" tree) to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological partition of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We infer phylogenetic trees from each matrix in both ML and Bayesian framework.
(4) We then compared the "missing-data" trees to the "best" tree.
We repeated step ~\ref{step:generate_matrix} to ~\ref{step:compare_topo} 51 (3$\times $17) times.}
\label{Fig_Outline}
\end{figure}

\begin{table}[ht]
\caption{Glossary}
\centering
\begin{tabular}{p{5cm} p{9cm}}
\hline
\hfill Term & Definition \\[0.5ex]
\hline
\hfill living taxa & taxa with both molecular and morphological data available \\[1.5ex]
\hfill fossil taxa & taxa with only morphological data available \\[1.5ex]
\hfill "complete" matrix & matrix with no missing data except for the molecular part of the fossil taxa \\[1.5ex]
\hfill "missing-data" matrix & matrix with various amount of missing data \\[1.5ex]
\hfill $M_{L}$ & missing living taxa in the morphological part of the matrix \\[1.5ex]
\hfill $M_{F}$ & missing data for the fossil taxa in morphological part of the matrix \\[1.5ex]
\hfill $M_{C}$ & missing morphological characters for both living and fossil taxa \\[1.5ex]
\hfill "true" tree & tree used to simulate the matrix \\[1.5ex]
\hfill "best" tree & tree inferred from the "complete" matrix \\[1.5ex]
\hfill "missing-data" tree & tree inferred the "missing-data" matrices \\[1.5ex]
\hfill RPBTC & Random pairwise Bayesian tree comparison \\[1.5ex]
\hfill "horizontal" sub-matrix & sub-matrix with morphological and molecular characters for living taxa \\[1.5ex]
\hfill "vertical" sub-matrix & sub-matrix with morphological data for both living and fossil taxa \\[1.5ex]
\hfill "corner" sub-matrix & sub-matrix with morphological data for living taxa \\[1.5ex]
\hfill "missing-data" sub-matrix & molecular part of the matrix for the fossil taxa (no data) \\
\end{tabular}
\label{Tab_glossary}
\end{table}

%---------------------------------------------
%       Generating the matrix
%---------------------------------------------

\subsection{Generating the matrix}
First we randomly generated a "true" tree of 50 taxa in R v3.0.2 \citep{R302} using the package diversitree v0.9-6 \citep{fitzjohndiversitree2012}.
We generated the tree using a Birth Death process by sampling the values of the speciation events ($\lambda$) and extinction events ($\mu$) from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}.
We implemented a rejection sampling algorithm to select only random trees with 25 living and 25 fossil taxa. %Justification needed?
We then added a taxa to the resulting Birth-Death tree as the outgroup of the tree.
The mean branch length of the tree was used to separate the outgroup from the rest of the taxa and the branch length leading to the outgroup was set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we created a molecular and a morphological matrix from the "true" tree.
The molecular matrix was inferred from the "true" tree using the package phyclust v0.1-14 \citep{chen2011}.
The matrix was made of 1000 characters sites for 51 taxa and generated using the seqgen algorithm \citep{ranbaut1997seqgen}.
We used the HKY model \citep{HKY85} with a random base frequencies and with the transition/transversion rate of 2 \citep{douadycomparison2003} as parameters for generating the matrix.
The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \citep{yangamong-site1996}.
We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal.
These parameters were selected to generate data with no special assumption about how the characters evolved as well as to reduce the computational time required if these parameters were estimated rather than defined (total computational time $>$ 65 CPU years).

We inferred the morphological matrix using the ape package v3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa.
We assigned the number of character states (either 2 or 3) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters.
These probabilities were selected using the overall distribution of characters states extracted from 100 published empirical morphological matrices (See \hyperref[supplementaries]{supplementaries}).
We then ran an independent discrete character simulation for each character using the "true" tree branch length and topology with the randomly selected number of states (2 or 3) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to an other \citep{Pagel22011994}.
This method allows us to have only two parameters per character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
We used a low evolution rate parameter (i.e. $\alpha$) in order to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014}.

All the molecular information for fossil taxa was replaced by missing data ("?").
Finally, we combined the morphological and molecular matrices obtained from the "true" tree.
Hereafter we call this the "complete" matrix: the matrix with no missing data except for the molecular data of the fossil taxa.

%---------------------------------------------
%       Removing data
%---------------------------------------------

\subsection{Removing data}
Once we obtained the "complete" matrix we modified it to get a set of matrices with missing data. %Ugly sentence
We randomly replaced data with "?" in the morphological part of the matrices according to the following parameters (Fig. ~\ref{Fig_RemoveData}):

\begin{enumerate}
\item{The proportion of living taxa with no morphological data ($M_{L}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological one.
Because of the increasing facility to sequence DNA for living taxa, the number of living taxa with molecular data is highly superior the the number of taxa with molecular and morphological data.
\item{The proportion of missing morphological data across all fossil taxa ($M_{F}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record. 
\item{The proportion of missing morphological characters across all taxa (living and fossil - $M_{C}$): 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa.
\end{enumerate}

\begin{figure}
\centering
\includegraphics[keepaspectratio=true]{Figures/TEM_Fig_missingData-BW.pdf}
\caption{Names of the different parts of the matrix.
A: Missing data parameters:
Missing living - The proportion of living taxa with no morphological data ($M_{L}$);
Missing fossil - The proportion of missing morphological data across all fossil taxa ($M_{F}$);
Missing character - The proportion of missing morphological characters across all taxa (living and fossil) ($M_{C}$).
B: Different parts of the matrix:
The "horizontal" sub-matrix (orange) contains molecular and morphological data for living taxa;
The "vertical" sub-matrix (blue) contains morphological data for living and fossil taxa;
The "corner" sub-matrix (orange and blue striped) contains morphological data for living taxa;
The "missing-data" sub-matrix (grey) is the molecular part of the fossil taxa and contains no data.}
\label{Fig_RemoveData}
\end{figure}

In practice, each parameter represent a different way of removing data in the morphological part of the matrix: $M_{L}$ removes a proportion of rows from the living taxa; $M_{F}$ removes a proportion of cells from the fossil taxa; and $M_{C}$ removes a proportion of columns across both living and fossil taxa (see Fig. ~\ref{Fig_RemoveData}).
Note that $M_{L}$ is different to $M_{F}$ not only because of the region of the matrix affected: for $M_{L}$, all the morphological data of a proportion of the living taxa is removed (i.e. removing rows), as for $M_{F}$, a proportion of data is removed across the whole of the morphological matrix for fossil taxa (i.e. removing cells).

We tested all parameters combinations resulting in 125 ($5^3$) matrices.
Because some parameter combinations introduce a lot of missing (e.g. $M_L$=75\%, $M_F$=75\% and $M_C$=75\%), some matrices contained fossil taxa without any data at all.
When this occurred we repeated the random deletion of characters until every taxa had at least 5\% data across the whole matrix.

%---------------------------------------------
%       Building phylogenies
%---------------------------------------------

\subsection{Building phylogenies}
From the resulting matrices we generated two types of trees, the "best" tree that is inferred from the "complete" matrix and the "missing-data" trees inferred from the 125 matrices with various amounts of missing data.
The "true" tree was used to generate the "complete" matrix and reflects the "true" evolutionary history in our simulations.
The "best" tree, on the other hand, is the best tree we can build using the state-of-the-art phylogenetic methods.
In real world situations, the "true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \citet{rozen2005}).
Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the "best" tree, rather than the "true" tree, as the "best" tree is generally what biologists have to work with.

\subsubsection{Maximum Likelihood}
The "best" tree and the "missing-data" trees were inferred using RAxML v8.0.20 \citep{Stamatakis21012014}.
For the molecular data, we used the GTR + $\Gamma_4$ model (\citet{tavare1986}; default GTRGAMMA in RAxML v8.0.20; \citet{Stamatakis21012014}) as a generalisation of the HKY + $\Gamma_4$ model \citep{HKY85} for the molecular data.
The GTR model can be seen as a generalisation of the HKY model (the 2 parameters from the HKY model are implicitly included in the 6 from GTR model - \citet{stamatakisa2008}).
For the morphological data, we used the implemented Markov \textit{k} state model \citep{lewisa2001} which is a generalisation of the JC69 model \citep{jc69} with \textit{k} $\geq$ 2 assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v8.0.20; \citet{Stamatakis21012014}).
We used the fast bootstrap algorithm and performed 1000 bootstraps per tree inference to assess the topological support. %The explanations of the algorithm part for the BS was proposed by Trevor
The bootstrap algorithm used in RAxML is the Lazy Sub-tree Rearrangement (LSR) which consists in pruning one sub-tree from the tree and subsequently reinserting it to al neighbouring branches \citep{stamatakisa2008}.
Sub-tree Pruning and Reinserting methods (SPR) have been demonstrated as being better than others (e.g. Nearest Neighbouring Interchange - NNI) in recovering good bootstrap values \citep{salamin2003}.

\subsubsection{Bayesian}
The "best" tree and the "missing-data" trees were inferred using MrBayes v3.2.2 \citep{Ronquist2012mrbayes}.
We partitioned the data to treat the molecular part as a non-codon DNA partition and the morphological part as a multi-state morphological partition.
The molecular evolutionary history was inferred using the HKY model with a transition/transversion ratio of two \citep{douadycomparison2003} and a gamma distribution for the rate variation with four distinct categories (HKY + $\Gamma_4$).
For the morphological data, we used the Markov \textit{k} state model \citep{lewisa2001}, with equal state frequency and a unique overall substitution rate ($\mu$) with four distinct rates categories (M\textit{k} + $\Gamma_4$).
We chose these models to be consistent with the parameters used to generate the "complete" matrix.

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 50$\times$$1^6$ generations.
We used the average standard deviation of split frequencies (ASDS) as a proxy to estimate the convergence of the chains and used a stop rule when the ASDS went below 0.01 \citep{Ronquist2012mrbayes}.
The effective sample size (ESS) was also checked on a random sub-sample of runs in each simulation to ensure that ESS $>>$ 200 \citep{drummond2006ess}.
For each run, we removed 25\% of the iterations as burn-in.
We used the following prior for each tree (see \hyperref[supplementaries]{supplementaries}):
\begin{enumerate}
\item
the "true" tree’s topology as a starting tree (with a starting value for each branch length of 1),
\item
an exponential prior on the shape of the gamma distribution of $\alpha$=0.5 for both partitions
\item
and a transition/transversion ratio prior of 2 sampled from a strong beta distribution ($\beta$(80,40)).
\end{enumerate}

We used these prior to speed up the Bayesian process.
These prior biased the way the Bayesian process calculated the branch length by giving non-random starting points and boundaries for the parameters estimation process, however, in this study, we focused on the effect of missing data on the topology and not on the branch length.
Even using these prior, it took 65 CPU years to build 51 sets of 125 Bayesian trees (8 core nodes 2.30GHz clock speed).

%---------------------------------------------
%       Comparing Topologies
%---------------------------------------------

\subsection{Comparing topologies}
We compared the topology of the "missing-data" trees inferred from the matrices with missing data to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$.
Note that we only investigate differences in topology and not in branch length because the aim of this study is to look at the effect of missing data on the topology of trees inferred from Total Evidence type matrices.
To compare the topology of the resulting trees, we used two metrics to assess number of conserved taxa and clades position using respectively the Triples \citep{dobson1975triplets} and the Robinson-Fould \citep{RF1981} distance.
We normalised the two metrics using the Normalised Tree Similarity index \citep{Bogdanowicz2012} to generalise our results for any \textit{n} number of taxa.
The three metrics are detailed below.

\subsubsection{Triples distance ($T_{x,y}$) \citep{dobson1975triplets}}
This metric measures the number of different sub-trees of three taxa between two given trees.
Each triplet can be written as $I_{ijk}$=(\textit{ijk}).
Where $I_{ijk}$ is equal to 0 if the the two triplets (\textit{ijk}) are the same in the two trees otherwise $I_{ijk}$ is equal to 1.
For any rooted binary tree there are only three possible combinations per triplets: ((\textit{j},\textit{k}),\textit{i});, ((\textit{i},\textit{k}),\textit{j}); and ((\textit{i},\textit{j}),\textit{k}); \citep{johnson1998}.
If the trees used are not fully binary, a fourth triplet combination is possible: (\textit{i},\textit{j},\textit{k});.
One can calculate $S_n$, the triplet distance between two trees as:
\begin{equation}
S_n = \sum_{ijk} I_{ijk}
\end{equation}
Where:
\begin{equation}
\sum_{ijk} = \binom{n}{4} = \frac{n!}{4!(n-4)!}
\end{equation}
And where \textit{n} is the number of taxa in both trees (modified from \citet{critchlowthe1996}).
If $S_n$=0, the trees are the same (i.e. no taxa as been displaced).
When $S_n$ = $\binom{n}{4}$, the trees are the most different possible (i.e. every taxa as been displaced).
This metric therefore illustrates the amount of displayed taxa.
It is less sensitive to the placement of individual taxa and to taxa of highly uncertain placement (e.g. fossil taxa) than the Robinson-Foulds metric \citep{critchlowthe1996,johnson1998,wiensmissing2003}.
Therefore we used this metric as a proxy to estimate the robustness of the tree to flying taxa (see \hyperref[supplementaries]{supplementaries}).

\subsubsection{Robinson-Fould distance (\textit{RF}) \citep{RF1981}}
This metric measures the number of shared clades among two trees and therefore illustrates the number of exactly conserved groups among the trees.
The Robinson-Fould distance (also called path difference) between two trees reflects the distance between the distributions of the tips among clades in the two trees \citep{RF1981} and can be expressed as following:
\begin{equation}
RF_{x,y} = N_{x} + N_{y} - 2C_{x,y}
\end{equation}
Where $C_{x,y}$ is the number of clades in common in the two trees.
The minimal value of \textit{C} is 1 if the two trees have the same \textit{n} taxa; the maximal value in \textit{C=n-2}.
This metric is more sensitive to taxa displacement than the Triples metric (i.e. if one taxa gets out of a clade, then the clades are no longer considered as similar - \citet{critchlowthe1996,johnson1998,wiensmissing2003}).
Therefore a low value will show a good clade conservation between two trees and a high value will show a bad recovery of common clades (see \hyperref[supplementaries]{supplementaries}).

>>>>

\subsection{Morphological characters states}
In order to obtain a realistic probabilistic value for of \textit{k} characters states for each simulated morphological character, we downloaded 100 random morphological characters (with more than 100 characters each) from TreeBASE database (http://treebase.org/) published between 1985 and 2013 and covering 19 taxomomic classess (Chordata, Arthropoda, Annelida, Angiosperm, Gymnosperm and Pteridophyta).
We selected a total of 22563 characters ranging from 2 to 10 states.
We calculated the proportion of characters with 2, 3, 4, 5, 6, 7, 8, 9 or 10 states.
We then sampeled 22563 \textit{k} values between 2 and 10 with the same proportion of characters from the empirical data.
We then used a simple t-test to check if our simulation was equal to the empirical data.
In this study, we only simulated characters with 2 or 3 states because of the high proportion of ordered characters ecountered on characters with more than 3 states and the difficulties of simulate biologicaly sensible ordered characters.

\begin{figure}
\centering
\includegraphics[keepaspectratio=true]{SupplementaryMaterial/Supp_Figures/TEM_Fig-AppendixCharacters.pdf}
\caption{Character states distribution in empirical matrices. %Title?
Characters states number distribution extracted from 100 random morphological matrices downloaded from RreeBase.}
\label{Fig_AppendixCharacters}
\end{figure}
